---
title       : Microbial Informatics
subtitle    : Lecture 17
date        : October 23, 2014
author      : Patrick D. Schloss, PhD (microbialinformatics.github.io)
job         : Department of Microbiology & Immunology
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      #
widgets     : mathjax       # {mathjax, quiz, bootstrap}
mode        : standalone    # {selfcontained, standalone, draft}
knit        : slidify::knit2slides

---

## Announcements
* Project 1 is due today
* Read ***The Art of R Programming*** (Chapters 6 and 7)

```{r eval=TRUE, echo=FALSE}

library(knitr)
opts_chunk$set(tidy = TRUE)

opts_chunk$set(hide = TRUE)
opts_chunk$set(eval = FALSE)
opts_chunk$set(cache = FALSE)
par(mar=c(4,5,0.25, 0.25))
```

---

## Review
* How to design a small function from pseudocode
* Additional details about vectors and matrices

---

## Learning objectives
* Lists
* Data frames
* Factors

---

## Lists
*	a data structure that can combine different data types (eg. numeric, vectors, strings, etc)
* useful for returning complex data structures from functions
* important for organizing data: imagine creating a list for each gene in a genome - gene name, coordinates, length, function, DNA sequence, translated sequence, a reference to experiments, etc...
* Could have a vector of lists... or a data frame...

---

## Creating a list

```{r, eval=TRUE, echo=TRUE}
ps <- list(name=c("Pat", "Schloss"), position="Associate Professor", time.at.um=5)
ps$name
ps$position
ps$time.at.um
```

---

## Working with lists

```{r, echo=TRUE, eval=TRUE}
str(ps)
length(ps)
names(ps)
```

---

## Working with lists

```{r, echo=TRUE, eval=TRUE}
unlist(ps)
is.vector(unlist(ps))
```

---
	
##	Accessing elements of a list
	
```{r, echo=TRUE, eval=TRUE}	
ps$position
ps[["position"]]
ps[[2]]
ps[2]
```

---
	
## Inserting things into a list

```{r, echo=TRUE, eval=TRUE}
ps$kids.names <- c("mary", "patrick", "joe", "john", "ruth", "jacob", "peter")
ps$kids[3]
ps[[4]][3]
ps$married <- TRUE
```	

---

```{r, eval=TRUE, echo=TRUE}
ps
```

---

## Deleting elements from a list

```{r, echo=TRUE, eval=TRUE}
ps$position <- NULL
ps
```

---

##	Applying functions to a list	

```{r, echo=TRUE, eval=TRUE}
data <- list(x=1:10, y=100:110)
lapply(data, median)	#returns a list
sapply(data, median)	#returns a vector
```

---	

## Working with function output

```{r list_struct, echo=TRUE, eval=FALSE}
x <- runif(100)
y <- c(rep("red", 50), rep("blue", 50))

t <- t.test(x~y)
str(t)
```

---

```{r list_struct, echo=FALSE, eval=TRUE}
```

---

```{r, eval=TRUE}
t$statistic
t$parameter
t$p.value
t$estimate
```

---

## Data frame
- like a matrix, but each column can have a different mode
-	We saw these in the first several classes when you were looking at the metadata file
-	A data frame is really a multidimensional list

---

## Creating a data frame

```{r, echo=TRUE, eval=TRUE}
name <- c("Schloss", "Young" , "Mobley", "SwansonJ")
rank <- c("Asst", "Assoc", "Chair", "Full")
cool <- c(10, 1, 9, 8)
directory <- data.frame(name, rank, cool)
directory
```

---

## Accessing columns of a dataframe

```{r, echo=TRUE, eval=TRUE}
directory[,1]
directory[,"name"]
```

---

## Accessing rows of a dataframe

```{r, echo=TRUE, eval=TRUE}
directory[1,]
directory[directory$name=="Schloss",]
rownames(directory) <- directory$name
directory["Schloss",]
```

---

## Factors
*	Vector that can contain only predefined values and is used to store categorical data.	

```{r, eval=TRUE, echo=TRUE}
x <- factor(c("a", "b", "b", "c", "d"))
x
str(x)
levels(x)
```

---

## Oops! have to define the values a priori

```{r, echo=TRUE, eval=TRUE}
x[2] <- "e"
x
```

---

## Define the values *a priori*

```{r, eval=TRUE, echo=TRUE}
x <- factor(c("a", "b", "b", "c", "d"), levels=c("a", "b", "c", "d", "e"))
x
levels(x)
table(x)
```

---

## Notice a difference?

```{r, eval=TRUE, echo=TRUE}
x <- factor(c("a", "b", "b", "c", "d"), levels=c("a", "e", "b", "c", "d"))
x
levels(x)
table(x)
```

---
	
##	Applying functions to factors: split
* splits a vector into groups

```{r, echo=TRUE, eval=TRUE}
set.seed(2)
ages <- sample(20:40, 20, replace=TRUE)
gender <- factor(sample(c("Female", "Male"), 20, replace=TRUE))
pol <- factor(sample(c("D", "R", "I"), 20, replace=TRUE))
split(ages, gender)
```

---

##	Applying functions to factors: cut

* generate factors based on pre-defined bins
	
```{r, echo=TRUE, eval=TRUE}
range <- seq(20,40,5)
segments <- cut(ages, range)	#(20,30]: 20 < x <= 30
table(segments)
```

---

##	Applying functions to factors: cut

* ... or let cut figure out the breaks for you

```{r, echo=TRUE, eval=TRUE}
segments <- cut(ages, 5)
table(segments)
```

---

##	Applying functions to factors: cut

* ... or make equally abundant groups

```{r, echo=TRUE, eval=TRUE}
range <- quantile(ages)
segments <- cut(ages, range)
table(segments)
```

--- .segue .dark

## Questions?
