--- 
title       : Microbial Informatics
subtitle    : Lecture 14
date        : October 10, 2014
author      : Patrick D. Schloss, PhD (microbialinformatics.github.io)
job         : Department of Microbiology & Immunology
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : mathjax       # {mathjax, quiz, bootstrap}
mode        : standalone    # {selfcontained, standalone, draft}
knit        : slidify::knit2slides

--- 

## Announcements
* Homework 3 is due today
* Have posted some guide lines to follow for Project 1
* No class on Tuesday (Fall break); Next Friday, the first hour of class will be a lecture
* Start programming discussion next Thursday!
* Read ***The Art of R Programming*** (Chapter 1: Getting Started)


```{r eval=TRUE, echo=FALSE}
library(knitr)
opts_chunk$set(hide = TRUE)
opts_chunk$set(eval = FALSE)
opts_chunk$set(tidy = TRUE)
opts_chunk$set(cache = FALSE)
par(mar=c(4,5,0.25, 0.25))
```

---

## Review



---

## Learning objectives

---

## Motivation  



---

## Motivation  

---

## Let's do an experiment...


```{r obs.data, echo=FALSE, fig.align='center', strip.white=TRUE, fig.height=6}
set.seed(1)
a <- rnorm(30, 6, 1)
b <- rnorm(30, 8, 1)
c <- rnorm(30, 6, 1)

stripchart(at=1, a, vertical=T, method="jitter", xlim=c(0,4), pch=19, ylim=c(0,max(a,b,c)), ylab="L. acidophilus (log CFUs/g)")
stripchart(at=2, b, vertical=T, method="jitter", pch=19, add=T)
stripchart(at=3, c, vertical=T, method="jitter", pch=19, add=T)

a.mean <- mean(a)
b.mean <- mean(b)
c.mean <- mean(c)

points(x=c(0.75, 1.25), y=c(a.mean, a.mean), type="l", lwd=3)
points(x=c(1.75, 2.25), y=c(b.mean, b.mean), type="l", lwd=3)
points(x=c(2.75, 3.25), y=c(c.mean, c.mean), type="l", lwd=3)

axis(1, at=seq(1:3), label=c("A", "B", "C"))

```


--- .segue .dark

## Questions?
