---
title       : Microbial Informatics
subtitle    : Lecture 19
date        : October 30, 2014
author      : Patrick D. Schloss, PhD (microbialinformatics.github.io)
job         : Department of Microbiology & Immunology
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      #
widgets     : mathjax       # {mathjax, quiz, bootstrap}
mode        : standalone    # {selfcontained, standalone, draft}
knit        : slidify::knit2slides

---

## Announcements
* Homework is due Friday
 - Note: for histogram problem, you should have the word names on the x-axis and their frequency in the y-axis (`hist` will not work...)
* We will not be meeting on Friday

```{r eval=TRUE, echo=FALSE}

library(knitr)
opts_chunk$set(tidy = TRUE)

opts_chunk$set(hide = TRUE)
opts_chunk$set(eval = FALSE)
opts_chunk$set(cache = FALSE)
par(mar=c(4,5,0.25, 0.25))
```

---

## Review
* Loops: `for`, `while`, `repeat` loops
* Loops are slow in R beacuase variables are copied, destroyed, and recreated each time a vector is modified
* Conditionals: `ifelse` and `if...else if...else`

---

## Learning objectives

* Be able to vectorize loops in R and see the improved speed
* Understand when to use Various flavors of apply
  - `apply`
  - `lapply`
  - `sapply`
  - `mapply`
  - `vapply`
  - `replicate`

---

## Simple vectorization

```{r, echo=TRUE, eval=TRUE, results='hold'}
z <- sum(1:10)

my.sum <- function(x){
	sum <- 0
	for(i in x){
		sum <- sum+i
	}
  return(sum)	
}

my.z <- my.sum(1:10)

z == my.z
```

---

## `apply`

* As we've seen before if we have a **matrix**, we can't easily perform functions on the rows...

```{r, eval=TRUE, echo=TRUE}
my.matrix <- matrix(runif(100), nrow=25, ncol=4)
mean(my.matrix)
```

* How do we get out the mean for each column?

---

## `apply`

* For the columns...
```{r, eval=TRUE, echo=TRUE}
apply(my.matrix, 2, sum)
```

* For the rows...
```{r, eval=TRUE, echo=TRUE}
apply(my.matrix, 1, sum)
```

---

## Something more complex...

* What if we want the sum of each column where each value is raised to some arbitrary power?

```{r}
power <- function(x, pow=2){
	value <- sum(x^pow)
	return(value)
}
```

* We could just loop across the columns and apply our `power` function:

```{r}
sum.pow <- rep(0, ncol(my.matrix))
for(c in 1:ncol(my.matrix)){
	sum.pow[c] <- power(my.matrix[,c], 3)
}
sum.pow
```

---

## Let's try the apply function...

```{r}
apply(my.matrix, 2, power)
```

* and a user defined power (`pow=3`)?

```{r, results='hold'}
apply(my.matrix, 2, power, pow=3)
```

---

## What's going on?

* “Returns a vector or array or list of values obtained by applying a function to margins of an array or matrix.”
* `apply` extracts the rows/columns, converts them to a vector
* The function is applied
* Output generated
* Again, the benefit is seen with larger datasets and more complex functions
* Remember that `apply` uses a matrix as input

---

## What if you want to use a list as your input?

```{r}
my.list <- list(popA = runif(20), popB = runif(40), popC = runif(25))
my.list
```

---

## Now I want to apply my `power` function to the elements in the list

```{r}
power(my.list)
```

* Ooops!

---

## `lapply` 

* “`lapply` returns a list of the same length as X, each element of which is the result of applying FUN to the corresponding element of X.”
* Give list (or vector) and `lapply` will perform the function over the elements within the list.

```{r}
lapply(my.list, power)
```

* Note that it returns a list.

---

## `sapply` 

* "`sapply` is a user-friendly version of lapply by default returning a vector or matrix if appropriate"
* If your output has the same length you can use the `sapply` option and you will get a vector as output

```{r}
sapply(my.list, power)
```

* Remember that you can give it a vector as well

---

## `vapply` 

* “`vapply` is similar to sapply, but has a pre-specified type of return value, so it can be safer (and sometimes faster) to use.”
* The syntax is a bit different: `vapply(X, FUN, FUN.VALUE, ...)` where `FUN.VALUE` is a vector with the name of the output from the function and its initial value

```{r}
vapply(my.list, power, c(value=0))
```

---

## `mapply`

* What if you have two vectors that you want to feed to a function?
* Say I want to raise each value to the power of it's position in the vector
* "`mapply` is a multivariate version of sapply. mapply applies FUN to the first elements of each (…) argument, the second elements, the third elements, and so on"

```{r}
my.vector <- runif(10)
my.powers <- 1:10
mapply(power, x=my.vector, pow=my.powers)
```

* Note that the function goes first and then the two vectors

---

## `replicate`

* Already saw this with the X^2 distribution homework problem...

```{r}
chi.sq <- function(k){
	rand.chisq <- sum(rnorm(k)^2)
	return(rand.chisq)
}

values <- replicate(1000, chi.sq(k=3))
head(values)
```

* Note that you need an actual function call for the second replicate arguement.

---

## Here's a problem...

* We have a table of relative abundances...

```{r}
relabund <- matrix(rep(runif(20000)), ncol=20, nrow=100)
relabund[5,] <- c(runif(10,0,0.4), runif(10, 0.3,0.7))
colnames(relabund) <- c(paste0("Lean", 1:10), paste0("Obese", 1:10))
rownames(relabund) <- paste0("Species", 1:100)
treatments <- c(rep("lean", 10), rep("obese", 10))

head(relabund)
```

---

## Here's a problem...

* Perform a wilcoxon test on each Species differentiating between lean and obese individuals

```{r, echo=FALSE}
test <- function(x, design){
	p <- wilcox.test(x~design)$p.value
	return(p)
}

p.values <- apply(relabund, 1, test, design=treatments)
padj.values <- p.adjust(p.values)
```
* Write R code, without the use of `for` loops that produces the following output:
  * `r rownames(relabund)[padj.values<0.05]` was significantly different between the two groups
  * In this statement, the "`r rownames(relabund)[padj.values<0.05]`" should be produced by r code
  * Be sure to correct for multiple comparisons!

--- .segue .dark

## Questions?
