---
title       : Microbial Informatics
subtitle    : Lecture 20
date        : November 4, 2014
author      : Patrick D. Schloss, PhD (microbialinformatics.github.io)
job         : Department of Microbiology & Immunology
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      #
widgets     : mathjax       # {mathjax, quiz, bootstrap}
mode        : standalone    # {selfcontained, standalone, draft}
knit        : slidify::knit2slides

---

## Announcements
* A new homework has been posted and is due on November 22nd
  * work with a partner
  * no more than one explicit loop
* Will have lab period on Friday

```{r eval=TRUE, echo=FALSE}

library(knitr)
opts_chunk$set(tidy = TRUE)

opts_chunk$set(hide = TRUE)
opts_chunk$set(eval = FALSE)
opts_chunk$set(cache = FALSE)
par(mar=c(4,5,0.25, 0.25))
```

---

## Review
* Loops are slow in R beacuase variables are copied, destroyed, and recreated each time a vector is modified
* the apply and replicate suite of commands allow you to minimize your use of explicit loops

---

## Here's a problem...

* We have a table of relative abundances...

```{r}
relabund <- matrix(rep(runif(20000)), ncol=20, nrow=100)
relabund[5,] <- c(runif(10,0,0.4), runif(10, 0.3,0.7))
colnames(relabund) <- c(paste0("Lean", 1:10), paste0("Obese", 1:10))
rownames(relabund) <- paste0("Species", 1:100)
treatments <- c(rep("lean", 10), rep("obese", 10))

head(relabund)
```

---

## Here's a problem...

* Perform a wilcoxon test on each Species differentiating between lean and obese individuals

```{r, echo=FALSE}
test <- function(x, design){
	p <- wilcox.test(x~design)$p.value
	return(p)
}

p.values <- apply(relabund, 1, test, design=treatments)
padj.values <- p.adjust(p.values)
```
* Write R code, without the use of `for` loops that produces the following output:
  * `r rownames(relabund)[padj.values<0.05]` was significantly different between the two groups
  * In this statement, the "`r rownames(relabund)[padj.values<0.05]`" should be produced by r code
  * Be sure to correct for multiple comparisons!

---

## Here's a problem...

* Perform a wilcoxon test on each Species differentiating between lean and obese individuals

```{r, echo=TRUE}
test <- function(x, design){
	p <- wilcox.test(x~design)$p.value
	return(p)
}

p.values <- apply(relabund, 1, test, design=treatments)
padj.values <- p.adjust(p.values)
```

---

## Here's a problem...
* Write R code, without the use of `for` loops that produces the following output:
  * ``r rownames(relabund)[padj.values<0.05]`` was significantly different between the two groups
  * `rownames(relabund)[padj.values<0.05]` was significantly different between the two groups
  * In this statement, the "``r rownames(relabund)[padj.values<0.05]``" should be produced by r code
  * In this statement, the "`rownames(relabund)[padj.values<0.05]`" should be produced by r code
* Be sure to correct for multiple comparisons!


---

## Learning objectives

* Understand how to get data in and out of R


---

## Integrating with other software

* **Problem:** We typically have large datasets and need a way to get that
information into the program.  Similarly, we generally have large
amounts of output that we can't just send to the screen.
			
* **Solution:**	We have seen a number of functions already for reading in data and
			in the very first homework you had a function to write out a table.
			R has a number of solutions to do both

---

## `readline`: read a single line from the screen

```
	> readline()
	ab de fg
	[1] "ab de fg"
	
	> readline("enter your name: ")
	Pat Schloss
	 
	> name <- readline("enter your name: ")
	Pat Schloss
	> name
	[1] "Pat Schloss"
```

---

## `scan`: reads in a vector from a file or keyboard

* Imagine a file that contains:

```
	123
	2 4 5
	53
```

```{r}
scan("file1.txt")
```	
	
* Note that it reads it in as a numeric vector


---
	
## `scan`: reads in a vector from a file or keyboard

* Now imagine a file that contains:

```
	123
	abcd
	2 4 5
	54
```

```{r}
scan("file2.txt")
```

---

## Try again...

```{r}
scan("file2.txt", what="")
```

* the default is to assume everything is a number

---

## Delimeters

* scan reads things into a vector assuming that any white space separates
	the vector values
	
```{r}
scan("file2.txt", what="", sep="\n")
```	

---

## Getting data from the keyboard

```
	> scan("")	#will read from the keyboard
	
	>scan("")
	1:  12 5 2 3
	5:  23 24
	6:
	Read 6 items
	[1] 12  5  2  3 23 24
```

* Note that to stop reading data in you give it a blank line


---

## `read.table`
* Reading in a table, every row has to have the same number of columns
*	Imagine a file that contains:

```
name	age sex
pat		36	M
john	22	M
susan	55	F
```

```{r}
read.table("table.txt", header=TRUE)
```

---

## `read.table`
* We'd like to get row names for our table

```{r}
read.table("table.txt", header=TRUE, row.names=1)
```

*	note that these commands would not have worked with the scan function since it is a mixture of text and numbers
* numerous other options for read.table including skipping rows, different delimeters, etc.

---

## You can read from the internet

```{r, tidy=TRUE}
data <- read.csv("http://archive.ics.uci.edu/ml/machine-learning-databases/echocardiogram/echocardiogram.data", header=F)
head(data)
```

* Note that `read.csv` is a special version of `read.table` for reading data talbe sthat are separated by commas 

---

##	Reading in lines of files

```{r, results='hold'}
c <- file("file2.txt", open="r")
readLines(c, n=1)
readLines(c, n=1)
readLines(c, n=1)
readLines(c, n=1)
readLines(c, n=1)
```	

---

## Be sure to `close` the file when you're done!

```{r, eval=FALSE}
close(c)
```

---
	
## `print`

* Writes to the screen

```{r}
x <- 1:3
print(x^2)
```

* This is useful for outputting data from within functions if you are trying to debug your code

---

## `cat`

* Write to the screen, must provide your own newline character

```{r}
cat(x, "abc", x^2)
cat(x, "abc", x^2, "\n")
```

* Not all implementations of R include a newline character with the cat function, so including the `\n` is useful to just to be safe

---

## `write.table`: structured writing to a file

```{r}
head(data)
colnames(data) <- LETTERS[1:ncol(data)]
rownames(data) <- 1:nrow(data)
write.table(file="ecg.data", data)
```

---

## File output:

```{r engine='bash'}
head ecg.data
```

* Again, many options for altering how the output looks

---

## Unstructured writing to a file

* `cat`:
```{r}	
	cat("abc\n", file="u")
	cat("def\n", file="u", append=T)
```

* `write`:
```{r}
	write(x, file="filename", sep= " ")
	write(x, file="filename", append=T, sep= "\t")
	write(x, "")
```

* Note that the `append=T` prevents the file from being written over

---

##	Basic file operations

* `file.info("data")`:	tells file size, creation time, etc
* `dir()`:	gives the directory contents for the specified dir
* `list.files()`:	same as dir()
* `file.exists()`:	tells whether a file exists, duh.
* `getwd()`:	tells you where you are in the file tree
* `setwd()`:	allows you to change the current directory

--- .segue .dark

## Questions?
