---
title       : Microbial Informatics
subtitle    : Lecture 22
date        : November 11, 2014
author      : Patrick D. Schloss, PhD (microbialinformatics.github.io)
job         : Department of Microbiology & Immunology
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      #
widgets     : mathjax       # {mathjax, quiz, bootstrap}
mode        : standalone    # {selfcontained, standalone, draft}
knit        : slidify::knit2slides

---

## Announcements
* A new homework has been posted and is due on November 22nd
  * work with a partner
  * no more than one explicit loop
* Will have lab period on Friday
* Read Chapters 11 in TAoRP for background material on what is discussed today

```{r eval=TRUE, echo=FALSE}

library(knitr)

opts_chunk$set(tidy = TRUE)
opts_chunk$set(hide = TRUE)
opts_chunk$set(eval = FALSE)
opts_chunk$set(cache = FALSE)
par(mar=c(4,5,0.25, 0.25))
```

---

## Review
* String manipulation
* Understand how to work with and manipulate character variables
* What do the following do?

```{r, eval=FALSE}
rev("abcdef")
paste(rev(unlist(strsplit("abcdef", ""))), collapse="")
```

---

## Learning objectives
* Finish discussing how to format output
* Using basic regular expressions

---

## Formatting text output with `sprintf`

```{r}
i <- 8
sprintf("the square of %d is %d", i, i^2)
sprintf("the square root of %d is %6.2f", i, sqrt(i))
sprintf("%d times 1e6 is %.3e", i, i * 1e6)
```

---

## Things to notice

* `%s` reserves the place for an string
* `%d` reserves the place for an integer
* `%f` reserves the place for an decimal number
* `%e` reserves the place for an number in scientific notation
* For `%f` and `%e` the format is `%m.n`. `n` indicates the number of values to the right of the decimal place to include and `m` indicates the total number of spaces to allot the string
* The output is a string
* Of course, you can do all of this in the text block of a knitr document

---

## Another useful way to format output to text

```{r, eval=FALSE}
format(x, trim = FALSE, digits = NULL, nsmall = 0L,
       justify = c("left", "right", "centre", "none"),
       width = NULL, na.encode = TRUE, scientific = NA,
       big.mark = "",   big.interval = 3L,
       small.mark = "", small.interval = 5L,
       decimal.mark = ".", zero.print = NULL,
       drop0trailing = FALSE, ...)`
```

* `x` is a number
* `trim` is whether to right justify numbers to a common width
* `digits` is the maximum number of significant digits
* `nsmall` is the minimum number of digits to the right of the decimal 

---

## Special vectors

```{r}
letters[1:5]
LETTERS[1:5]
```

---

## Special functions to manipulate character vectors

```{r}
toupper(letters[1:5])
tolower(LETTERS[1:5])
```

---

## Searching for sub-strings

*	`grep(pattern, x)` - "Get Regular ExPression"

```{r}
grep("A", c("ATA", "CCC", "CTA"))
grep("G", c("ATA", "CCC", "CTA"))
```

* `grep` tells you which elements of x have the pattern

---

## Searching for sub-strings

*	`grepl(pattern, x)` - "Get Regular ExPression" logic-based

```{r}
grep("A", c("ATA", "CCC", "CTA"))
grep("G", c("ATA", "CCC", "CTA"))
```

* `grepl` tells you whether or not each element of x ahs the pattern

---

## Where in the string does the pattern match?

*	`regexpr(pattern, x)` - identify the starting position of the pattern in x

```{r}
regexpr("ATG", "CTATGCATGC")
```

---

## But what if the pattern appears multiple times?

*	`gregexpr(pattern, x)` - 	identify the starting position of the pattern throughout x ~ global

```{r}
gregexpr("ATG", "CTATGCATGC")
```

---

## Find and replace

*	`sub(pattern, replacement, x)` - SUBstitute the PATTERN with the REPLACEMENT

```{r}
sub("T", "U", "CTATGCATGC")
```

* Notice anything odd? What command do you think would fix this?

---

## A global find and replace

*	`gsub(pattern, replacement, x)` - replaces all instances

```{r}
gsub("T", "U", "CTATGCATGC")
```

---

## How would you get the complement of a DNA sequence using `gsub`?

```{r}
dna <- "CTATGCATGC"
```
```{r, echo=FALSE}
dna <- gsub("A", "X", dna)
dna <- gsub("T", "A", dna)
dna <- gsub("X", "T", dna)
dna <- gsub("C", "X", dna)
dna <- gsub("G", "C", dna)
dna <- gsub("X", "G", dna)
```

The complement should be `r dna`

---

## A specialized find and replace

* `chartr(findChars, replaceChars, x)`

```{r}
chartr("T", "A", "CTATGCATGC")
chartr("TA", "AT", "CTATGCATGC")
chartr("TACG", "ATGC", "CTATGCATGC")
```

---

## Getting a bit more sophisticated

*	Ignoring the case of the text
	
```{r}
grep("g", "ATGCATGC")
grep("g", "ATGCATGC", ignore.case=T)
```

---

## Getting a bit more sophisticated

*	Location
	- ^	-	Beginning of a string
	- $	-	End of a string
	
```{r, eval=FALSE}
grep("^ATG", vectorOfGenes)
grep("TAA$", vectorOfGenes)
```

---

## Combining searches

*	|	-	Matches with either the expression before or after the pipe

```{r, eval=FALSE}
grep("TAA$|TAG$", vectorOfGenes)
```

---

## Let's revisit the metadata file

```{r}
metadata <- read.table(file="wild.metadata.txt", header=T)
head(metadata)
```

* The `Date` column is the date that the mice were captured in `M_DD` format. Can you convert this column into "Month Day, Year" format? Assume the year was 2011.

--- .segue .dark

## Questions?
